title: "Spotify Listening Analysis"
author: "Ryan Carroll"
subtitle: "https://github.com/rcarroll9389" 
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    runtime: shiny
resource_files:
- IMG_7628.jpeg
---

```{r include = FALSE, warning = FALSE}
#Load Libraries -----------
library(tidyverse)
library(jsonlite)
library(data.table)
library(lubridate)
library(ggplot2)
library(shiny)
library(DT)
library(shinydashboard)
```

```{r include = FALSE}
# drive <- "/Users/ryancarroll/Desktop/Spotify Data/Spotify Extended Streaming History"
# 
# files <- dir(drive, pattern = "*.json")
# 
# #Bringing in 20 separate json files with information on my streaming history
# spotifiles <- files %>%
#   map_df(~fromJSON(file.path(drive, .), flatten = TRUE))
# # 
 #write_csv(spotifiles, "/Users/ryancarroll/Desktop/Spotify Data/spotifiles.csv") 

spotifiles <- read_csv("spotifiles.csv")
```

# Every year, Spotify does an end of year review that gives you an idea of the top music you listen to, along with other interesting metrics such as, top songs, artists, genres, and where you rank among listening hours to the rest of the world.

# I have listened to a lot of music over the years. I am regularly in the top 95% or more of all listeners worldwide!

# According to the latest Spotify Wrapped, I was in the top 1% of all listeners in the world, listening to roughly 96,000 minutes of music!

# I'd say maybe I need a hobby, but I always have music playing in the background of everything I do. Whether I'm working out, skiing, biking, coding, or just hanging out. I always have something playing! I want to explore my Spotify data, and see what insights I can find from the past 10 years of data. The insights from Spotify Wrapped are great, but I want to expand them a little more, and create an interactive dashboard so I can see who exactly I spend my time listening to.

# Let's first explore the data set to see what features our data set holds.

```{r echo = FALSE}
#Glimpse one of the datasets to understand the data in each element 
  glimpse(spotifiles)
```

```{r include = FALSE}
#I am only worried about music files, so am going to remove all podcast and audiobooks I have listened too. 

spotifiles_clean <- spotifiles %>% 
  filter(!is.na(master_metadata_track_name)) %>%
  select(-contains(c("audiobook", "episode"))) 

spotifiles_clean2 <- spotifiles_clean %>%
  #Clean the Time Stamp (ts) column to be a date time
  mutate(ts_clean = lubridate::ymd_hms(ts)) %>%
  #separate date and the time component (which is in military time)
  separate(ts_clean, c("date", "time_military"), sep = " ") %>%
  mutate(date = ymd(date)) %>%
  separate(date, c("year", "month", "day"), sep = "-", remove = FALSE) %>%
  unite("month_year", month, year, sep = "-", remove = FALSE) %>%
  #convert miliseconds to seconds
  mutate(seconds_played = (ms_played * 0.001), 
         minutes_played = round(seconds_played / 60, 2)) %>%
  filter(!is.na(date))


# I want to see how many minutes I have listened to music on Spotify
spotify_minutes_listened <- spotifiles_clean2 %>%
  summarize(minutes_listened = sum(minutes_played), hours_listened = round(minutes_listened / 60)) %>%
  print()
```

# Wow! After cleaning the data, I was able to calculate that I have listened to `r format(spotify_minutes_listened$minutes_listened)` minutes (`r format(spotify_minutes_listened$hours_listened)` hours) since 2014. That is a lot of time listening to music over the last 10 years!

```{r include = FALSE}
time_listened <- spotifiles_clean2 %>%
  group_by(year) %>%
  mutate(minutes_per_year = round(sum(minutes_played))) %>%
  ungroup() %>%
  group_by(month, year) %>%
  mutate(minutes_per_month = sum(minutes_played)) %>%
  ungroup()

#View newly made column by year
time_listened_per_year <- unique(time_listened[c("year", "minutes_per_year")])

```

# Below is a line chart showing the number of minutes spent listening on Spotify since 2014. Looks like I have steadily increased my music listening over the past decade!

```{r echo = FALSE}
ggplot(time_listened, aes(x = year, y = minutes_per_year, group = 1)) +
  geom_line(color = "#1ed760") + 
  geom_point(color = "#1ed760") + 
  labs(x = "",y = "Minutes", title ="Total Minutes Listened to by Year")
```

# However, 2014 looks pretty low compared to the rest of the time series. After some exploration, it appears that 2014 only includes November and December data. Since this year and 2025 are incomplete, I exluded them from this analysis.

```{r echo = FALSE, include = FALSE}
# Is 2014 low because I didn't listen as often or is something else going on?
low_years <- time_listened %>%
  filter(year == 2014) %>%
  group_by(month) %>% 
  distinct(month, year)

table(low_years)

# Interesting, so only November and December are included for 2014. Let's look at all of the years and see if any other are incomplete. 
```

The table below shows months across all years. Months with a 0 indicate that the month is not included in the data set.

```{r echo = FALSE}
time_listened_check_years <- time_listened %>% 
#Let's double check all other years have full data (2023 will be incomplete as this was downloaded part way through the year) 
group_by(month, year) %>% 
  select(month, year) %>%
  distinct() 

table(time_listened_check_years)

# Looks like all other years are complete up until 2025. Will remove 2025 data as well as 2014
time_listened_cleaned_years <- time_listened %>%
  filter(year >= 2015 & year <2025) 

ggplot(time_listened_cleaned_years, aes(x = year, y = minutes_per_year, group = 1)) +
  geom_line(color = "#1ed760") + 
  geom_point(color = "#1ed760") + 
  labs(x = "", y = "Minutes", title ="Total Minutes Listened to by Year", caption = "Total Minutes by Year, 2015-2024")

```

```{r include=FALSE}

artist_analysis <- time_listened_cleaned_years %>%
  #filter(year == 2024) %>%
  group_by(year, master_metadata_album_artist_name) %>% 
  #Counting the total number of rows by artist. This is equal to the total number of songs listened to by each artist in the dataset
  summarise(top_value = n()) %>% 
  slice_max(top_value, n = 10) %>%
  mutate(indicator = "Top Artists") %>%
  rename("top_artist_song" = "master_metadata_album_artist_name")

song_analysis <- time_listened_cleaned_years %>%
  group_by(year, master_metadata_track_name) %>% 
  #Counting the total number of rows by artist. This is equal to the total number of songs listened to by each artist in the dataset
  summarise(top_value = n()) %>% 
  slice_max(top_value, n = 10)  %>%
  mutate(indicator = "Top Songs") %>%
  rename("top_artist_song" = "master_metadata_track_name" )

top_artist_song <- bind_rows(artist_analysis, song_analysis)

artist_song_counts <- time_listened_cleaned_years %>%
  select(year, master_metadata_album_artist_name, master_metadata_track_name, minutes_per_year) %>%
  pivot_longer(values_to = "value", names_to = "kpi", cols = c("master_metadata_album_artist_name", "master_metadata_track_name")) %>%
 group_by(year, kpi) %>%
  mutate(value = n_distinct(value)) %>%
  distinct()%>%
  ungroup()

```

# Please follow this link to view two interactive dashboards!
# https://rcarroll9389.shinyapps.io/Spotify_Analysis_Dashboards/



Thank you for checking out my dashboard! I am always looking for new data and artists to listen to so please send me suggestions at [rcarroll9389\@gmail.com](mailto:rcarroll9389@gmail.com){.email}

This will be an evolving analysis, and I will continue to add to explore the data further. The code behind this dashboard can be found on my Github at: <https://github.com/rcarroll9389>

